---
title: "Exploratory Analysis of UFO Sightings from NUFORC Dataset"
author: "Shreyansh Misra, Lawrence Meng"
date: "2024-11-20"
output:
  pdf_document:
    number_sections: true
---
```{r echo=FALSE, include=FALSE}
library(cluster)
library(ggplot2)
library(caret)
library(mclust)
library(forecast)
library(tseries)
library(knitr)
library(dplyr)
```
The NUFORC Databank is the largest independently collected set of unidentified flying object (UFO) and unidentified aerial phenomenon (UAP) sighting reports available on the internet. This project explores patterns and trends in UFO sightings to address the following questions:
1. **Where are UFOs most likely to be sighted?** Are they concentrated in specific countries, near landmarks, or certain distances from the equator?  
2. **When are UFOs most likely to be sighted?** Are sightings tied to specific seasons, holidays, or days of the week?  
3. **What are the most common UFO descriptions?** What shapes, patterns, and accounts are commonly reported?  

## Data
### Overview
The dataset comprises UFO sighting reports, with variables detailing locations, times, descriptions, and other relevant information.

#### Variables
- **Datetime**: When the sighting occurred.
- **City/State/Country**: The geographical location of the sighting.
- **Shape**: Reported shape of the UFO (e.g., "disk," "sphere", "fireball", "light", etc.).
- **Duration (seconds)**: Length of the sighting.
- **Latitude/Longitude**: Geographic coordinates of the sighting.
- **Comments**: Textual eyewitness accounts.

### Variables of Interest
- **Datetime**
- **City/State/Country**
- **Shape**
- **Latitude/Longitude**
Analysis of these variables will help us answer the question: When and where do different kinds of UFOs, in terms of shape, appear?

```{r}
# Load data
ufo.data <- read.csv("scrubbed.csv")

# Replace empty strings with "Unknown" in the state column
ufo.data$state <- ifelse(ufo.data$state == "", "Unknown", ufo.data$state)

# Recalculate state frequencies
state.freq <- table(ufo.data$state)

# Exclude "Unknown" from the top 10
state.freq <- state.freq[names(state.freq) != "Unknown"]
top_states <- sort(state.freq, decreasing = TRUE)[1:10]

# Bar plot for top states
barplot(top_states,
        main = "Top 10 US States by UFO Sightings (Excluding Unknown)",
        col = "blue",
        xlab = "State",
        ylab = "Number of Sightings")

```
```{r}
# Converting the datetime column to a easy-to-read date format
ufo.data$datetime <- as.Date(ufo.data$datetime, format = "%m/%d/%Y")

# Checking the range of dates
range(ufo.data$datetime)

#Filtering the range of dates to have recent data
ufo.filtered <- ufo.data %>%
  filter(datetime >= as.Date("1990-01-01") & datetime <= as.Date("2014-05-01"))

# Verifying the filtered date range
range(ufo.filtered$datetime)
``` 

```{r}
# Creating a year-month column
ufo.filtered$year_month <- format(ufo.filtered$datetime, "%Y-%m")

# Aggregating sightings by year-month
ufo.monthly <- ufo.filtered %>%
  group_by(year_month) %>%
  summarize(sightings = n())

# Converting year-month to Date type for plotting
ufo.monthly$date <- as.Date(paste0(ufo.monthly$year_month, "-01"))


```

```{r}
# Create a time series object
ufo.ts <- ts(
  ufo.monthly$sightings,
  frequency = 12, # Monthly frequency
  start = c(as.numeric(format(min(ufo.monthly$date), "%Y")),
            as.numeric(format(min(ufo.monthly$date), "%m")))
)

# Plotting the time series
autoplot(ufo.ts) +
  labs(title = "Monthly UFO Sightings (1990-2014)", x = "Year", y = "Sightings") +
  theme_minimal()



```

```{r}
# Decomposing the time series
ufo.decomp <- decompose(ufo.ts)

# Plotting the decomposed time series
autoplot(ufo.decomp) +
  labs(title = "Decomposition of Monthly UFO Sightings") +
  theme_minimal()



```

```{r}
# Augmented Dickey-Fuller Test
adf.test(ufo.ts)

# KPSS Test
kpss.test(ufo.ts)
```
For the ADF test, the p-value is very small, so the null hypothesis is rejected, and the time series is stationary.

For the KPSS test, the p-value is very small, so the null hypothesis is rejected, and the time series is not stationary.

These two are contradicting, so we apply differencing.

```{r}
# Differencing the time series
ndiffs(ufo.ts)
ufo.ts.diff <- diff(ufo.ts, differences = 1)

# ADF test
adf.test(ufo.ts.diff)

# KPSS test
kpss.test(ufo.ts.diff)

```
Now, the time series is stationary. 



